---
title: 'CS6190: Probabilistic Modeling Homework 2
Bayesian Networks'
author: "Gopal Menon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{mathtools}
- \usepackage{amsmath}
- \usepackage{dirtytalk}
- \DeclareMathOperator{\Unif}{Unif}
- \DeclareMathOperator{\E}{E}
- \DeclareMathOperator{\Var}{Var}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\hrule

\section*{Written Part}

\begin{enumerate}

\item The total probability of the network is given by

$$ 
\begin{aligned} 
p(Beysian\;Network)&=p(income)p(smoke|income)p(exercise|income)p(bmi|income,exercise)\\&\times p(blood\;pressure|exercise, income, smoking)p(cholesterol| exercise, income, smoking) \\ &\times p(diabetes | bmi)p(stroke | bmi, bp, cholesterol)p(attack | bmi, bp, cholesterol)\\ &\times p(angina | bmi, bp, cholesterol)
\end{aligned}
$$
The total probability distribution requires the product of $10$ random variables. If the probabilities are multiplied as shown above, the probability table will have $10$ variables. Number of allowed states for the random variables are:

\end{enumerate}

```{r init, echo=FALSE, results='asis'}
source("BRFSS Bayes Net.R")
load_libraries()
survey_variables = c("income", "exercise", "smoke", "bmi", "bp", "cholesterol", "angina", 
                            "stroke", "attack", "diabetes")
beysian_network_vars_states = data.frame(survey_variables, survey_variables_states)
names(beysian_network_vars_states) = c("Survey Variable", "Number of states")
print(kable(beysian_network_vars_states, caption = "CDC Survey Variable states"))
```

The maximum number of rows in the probability distribution table for the above variables will be the product of the number of allowed states - $8\times2\times2\times4\times4\times2\times2\times2\times2\times4=32768$. This is the number of probabilities that are needed to store the full joint distribution.

However it is easier and cheaper in terms of memory requirements to store probabilities corresponding to each node in the Bayes Net. For example the number of probabilities that need to be stored for the node exercise corresponds to $p(exercise|income)$ which will have $2\times 8=16$ different probabilities that need to be stored. The following table shows the number of probabilities that need to be stored at the level of each node.

```{r q1contd, echo=FALSE, results='asis'}
node_probabilities = c(probabilities_in_node(c("income")), 
                       probabilities_in_node(c("exercise", "income")), 
                       probabilities_in_node(c("smoke", "income")), 
                       probabilities_in_node(c("bmi", "exercise", "income")), 
                       probabilities_in_node(c("bp", "exercise", "income", "smoke")), 
                       probabilities_in_node(c("cholesterol", "exercise", "income", "smoke")), 
                       probabilities_in_node(c("angina", "bmi", "bp", "cholesterol")), 
                       probabilities_in_node(c("stroke", "bmi", "bp", "cholesterol")), 
                       probabilities_in_node(c("attack", "bmi", "bp", "cholesterol")),
                       probabilities_in_node(c("diabetes", "bmi")))

variables_column = survey_variables
variables_column[length(variables_column) + 1] = "Total"
node_probabilities[length(node_probabilities) + 1] = sum(node_probabilities)

beysian_network_vars_probabilities = data.frame(variables_column, 
                                                node_probabilities)
names(beysian_network_vars_probabilities) = c("Survey Variable", "Number of Probabilities")
print(kable(beysian_network_vars_probabilities, caption = "Number of Probabilities to be stored at node level"))
```

So it is more efficient to store probabilities at the level of a node and based on the query, marginalize variables that we are not interested in after observing the variables that are given (on the right side of the conditional probability). 

\begin{enumerate}
  \item[2.] For each of the four health outcomes (diabetes, stroke, heart attack, angina) 
  \begin{enumerate}
    \item Probability of the outcome if I have bad habits (smoke and don’t exercise)? How about if I have good habits (don’t smoke and do exercise)?
  \end{enumerate}
\end{enumerate}

```{r q1q2, echo=FALSE, results='asis'}

## 1.
# Create the Bayesian network to analyze the survey results. It will be the product of 
# all the probabilities corresponding to the full joint distribution. 
risk_factors_bayes_net_q1 = create_bayes_net()

## 2.
# For each of the four health outcomes (diabetes, stroke, heart attack, angina), answer the following by 
# querying the network

## (a) What is the probability of the outcome corresponding to bad and good habits
impact_of_habits_on_outcomes(bayes_net=risk_factors_bayes_net_q1)
```

\begin{enumerate}

  \item[(b)] What is the probability of the outcome if I have poor health (high blood pressure, high cholesterol, and overweight)? What if I have good health (low blood pressure, low cholesterol, and normal weight)?

\end{enumerate}

```{r q2contd, echo=FALSE, results='asis'}

## (b) What is the probability of the outcome corresponding to bad and good health
impact_of_health_on_outcomes(bayes_net=risk_factors_bayes_net_q1)
```

\begin{enumerate}

  \item[3.] Evaluate the effect a person’s income has on their probability of having one of the four health outcomes (diabetes, stroke, heart attack, angina). For each of these four outcomes, plot their probability given income status (your horizontal axis should be $i = 1,2,\ldots,8$, and your vertical axis should be $P(y = 1|income = i)$, where $y$ is the outcome). What can you conclude?

\end{enumerate}

```{r q3, echo=FALSE, results='asis'}
# Outcome probabilities by income level
diabetes_probability_by_income = numeric()
stroke_probability_by_income = numeric()
attack_probability_by_income = numeric()
angina_probability_by_income = numeric()

# For each income level
for (income_level in 1:8) {
  
  # Find the probability of a bad outcome
  outcome_tables = get_health_outcomes(bayes_net=risk_factors_bayes_net_q1, 
                                       health_outcomes=health_outcomes, 
                                       observe_variables=c("income"), 
                                       observe_values=c(income_level))
  
  # Store diabetes outcome probability
  diabetes_factor_table = outcome_tables[[which(health_outcomes == "diabetes")]]
  probability_value = diabetes_factor_table[which(diabetes_factor_table[["diabetes"]] == diabetes_values[1]), 1]
  diabetes_probability_by_income[income_level] = probability_value
  
  # Stroke stroke outcome probability
  stroke_factor_table = outcome_tables[[which(health_outcomes == "stroke")]]
  probability_value = stroke_factor_table[which(stroke_factor_table[["stroke"]] == stroke_values[1]), 1]
  stroke_probability_by_income[income_level] = probability_value
  
  # Store heart attack outcome probability
  attack_factor_table = outcome_tables[[which(health_outcomes == "attack")]]
  probability_value = attack_factor_table[which(attack_factor_table[["attack"]] == attack_values[1]), 1]
  attack_probability_by_income[income_level] = probability_value
  
  # Store angina outcome probability
  angina_factor_table = outcome_tables[[which(health_outcomes == "angina")]]
  probability_value = angina_factor_table[which(angina_factor_table[["angina"]] == angina_values[1]), 1]
  angina_probability_by_income[income_level] = probability_value

}

# Plot the health outcome probabilities by income level
par(mar=c(8,4,1,6))
plot(1:8, diabetes_probability_by_income, main="Health Outcome Probability by Income Level", type='l', col="red", xlab="", ylab="Outcome Probability", xaxt="n", ylim=c(0.03,0.16), lwd=3)
lines(1:8, stroke_probability_by_income, col="blue", lwd=3)
lines(1:8, attack_probability_by_income, col="green", lwd=3)
lines(1:8, angina_probability_by_income, col="magenta", lwd=3)
axis(side=1, at=1:8, labels=c("< $10,000","$10,000 - $15,000","$15,000, - $20,000","$20,000 - $25,000","$25,000 - $35,000","$35,000 - $50,000","$50,000 - $75,000","> $75,000"), las=2)
legend("topright", xpd=TRUE, text.width=0.9, inset=c(-0.2,0), legend=c("diabetes", "angina", "attack", "stroke"), col = c("red", "magenta", "green", "blue"), lwd=3, bg="white")
```

It looks like overall the probability of a bad health outcome reduces with increasing income. However, it looks like the probability first increases from the lowest to the next highest income and then reduces.

\begin{enumerate}

  \item[4.] Notice there are no links in the graph between the habits (smoking and exercise) and the outcomes. What assumption is this making about the effects of smoking and exercise on health problems? Let’s test the validity of these assumptions. Create a second Bayesian network as above, but add edges from smoking to each of the four outcomes and edges from exercise to each of the four outcomes. Now redo the queries in Question 2. What was the effect, and do you think the assumptions of the first graph were valid or not?

I think the assumption made in the graph is that the habits (smoking and exercise) have an impact on the health (bmi, bp and cholesterol) of the person and thus indirectly impact the outcomes (diabetes, stroke, attack and angina).
\end{enumerate}

```{r q4, echo=FALSE, results='asis'}
# Make habits impact outcomes
risk_factors_bayes_net_q4 = make_habits_impact_outcomes()

## (a) What is the probability of the outcome corresponding to bad and good habits
impact_of_habits_on_outcomes(bayes_net=risk_factors_bayes_net_q4)

## (b) What is the probability of the outcome corresponding to bad and good health
impact_of_health_on_outcomes(bayes_net=risk_factors_bayes_net_q4)
```

Comparing tables for health outcomes based on habits, we can see that when we have links from habits to outcomes,the probability for the outcome is higher for bad habits and lower for good habits in the latter case. This suggests that habits do directly impact the outcomes and so the second graph is a more accurate representation of the model. Its not that the first graph was invalid, it was less accurate.

Comparing tables for health outcomes based on health, we can see that when we have links from health to outcomes,the probability for the outcome is higher for bad health and lower for good health in the latter case. Again, this suggests that habits do directly impact the outcomes and so the second graph is a more accurate representation of the model.

\begin{enumerate}

\item[5.] Also notice there are no edges between the four outcomes. What assumption is this making about the interactions between health problems? Make a third network, starting from the network in Question 4, but adding an edge from diabetes to stroke. For both networks, evaluate the following probabilities:

P(stroke = 1|diabetes = 1) and P(stroke = 1|diabetes = 3)

Again, what was the effect, and was the assumption about the interaction between diabetes
and stroke valid?

\end{enumerate}

```{r q5, echo=FALSE, results='asis'}
# Make diabetes impact stroke outcome
risk_factors_bayes_net_q5 = make_diabetes_impact_stroke_outcome()

#Print the impact
impact_of_diabetes_on_stroke(bayes_net=risk_factors_bayes_net_q4, FALSE) 
impact_of_diabetes_on_stroke(bayes_net=risk_factors_bayes_net_q5, TRUE) 
```

The assumption being made when there are no edges between the four outcome is that only the health conditions (and the habits when there is a link from them to the outcomes), affect the outcomes and that the presence of one outcome does not impact the probability of another outcome.

When there is no edge in the bayes net from diabetes to stroke, the probability of stroke does not vary much whether diabetes is present or not present. However when there is an edge, probability of stroke is much higher when diabetes is present compared to when it is not. So this suggests that there is an impact of diabetes on the probability of having a stroke. So this suggests that the graph with the edge from diabetes to stroke is a more accurate representation of the model.

\begin{enumerate}

\item[6.] Finally, make sure that your code runs correctly on all of the examples in BayesNetExamples.r. Your code will be graded for correctness on these also.

\end{enumerate}

```{r q6, echo=FALSE, results='asis'}
source("BayesNetworkExamples.r")

cbt8.30 = infer(carNet, c("battery", "fuel"), NULL, NULL)     ## (8.30)
cbt8.31 = infer(carNet, c("battery"), "fuel", 0)              ## (8.31)
cbt8.32 = infer(carNet, c("battery"), "gauge", 0)             ## (8.32)
cbt8.33 = infer(carNet, NULL, c("gauge", "battery"), c(0, 0)) ## (8.33)

print(kable(cbt8.30, caption = "Bishop 8.30 - p(G = 0)"))  
print(kable(cbt8.31, caption = "Bishop 8.31 - p(G = 0|F = 0)"))  
print(kable(cbt8.32, caption = "Bishop 8.32 - p(F = 0|G = 0)"))  
print(kable(cbt8.33, caption = "Bishop 8.33 - p(F = 0|G = 0, B = 0)"))  

kmt1 = infer(grassNet, c("cloudy", "rain"), "wet", "T")                                      ## Pr(S|W=1)
kmt2 = infer(grassNet, c("cloudy", "sprinkler"), "wet", "T")                                 ## Pr(R|W=1)
kmt3 = infer(grassNet, c("cloudy", "sprinkler", "rain"), NULL, NULL)                         ## Pr(W)
kmt4 = infer(grassNet, c("cloudy"), c("wet", "rain"), c("T", "T"))                           ## Pr(S|W=1,R=1)

print(kable(kmt1, caption = "Kevin Murphy - Pr(S=1|W=1)"))  
print(kable(kmt2, caption = "Kevin Murphy - Pr(R=1|W=1)"))  
print(kable(kmt3, caption = "Kevin Murphy - Pr(W=1)"))  
print(kable(kmt4, caption = "Kevin Murphy - Pr(S=1|W=1,R=1)"))  
```
