---
title: 'CS6190: Probabilistic Modeling Homework 2
Bayesian Networks'
author: "Gopal Menon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{mathtools}
- \usepackage{amsmath}
- \usepackage{dirtytalk}
- \DeclareMathOperator{\Unif}{Unif}
- \DeclareMathOperator{\E}{E}
- \DeclareMathOperator{\Var}{Var}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\hrule

\section*{Written Part}

\begin{enumerate}

\item The total probability of the network is given by
\textbf{Hint:} Start with the identity $\int p(x) dx = 1$, and take the derivative with respect to $\eta$.

$$ 
\begin{aligned} 
p(Beysian\;Network)&=p(income)p(smoke|income)p(exercise|income)p(bmi|income,exercise)\\&\times p(blood\;pressure|exercise, income, smoking)p(cholesterol| exercise, income, smoking) \\ &\times p(diabetes | bmi)p(stroke | bmi, bp, cholesterol)p(attack | bmi, bp, cholesterol)\\ &\times p(angina | bmi, bp, cholesterol)
\end{aligned}
$$
The total probability distribution requires the product of $10$ random variables. If the probabilities are multiplied as shown above, the probability table will have $10$ variables. Number of allowed states for the random variables are:

\end{enumerate}

```{r init, echo=FALSE, results='asis'}
source("BRFSS Bayes Net.R")
load_libraries()
survey_variables = c("income", "exercise", "smoke", "bmi", "bp", "cholesterol", 
                     "angina", "stroke", "attack", "diabetes")
survey_variables_states = c(8, 2, 2, 4, 4, 2, 2, 2, 2, 4)
beysian_network_vars_states = data.frame(survey_variables, survey_variables_states)
names(beysian_network_vars_states) = c("Survey Variable", "Number of states")
print(kable(beysian_network_vars_states, caption = "CDC Survey Variable states"))
```

The maximum number of rows in the probability distribution table for the above variables will be the product of the number of allowed states - $8\times2\times2\times4\times4\times2\times2\times2\times2\times4=32768$. This is the number of probabilities that are needed to store the full joint distribution.

\begin{enumerate}
  \item[2.] For each of the four health outcomes (diabetes, stroke, heart attack, angina) 
  \begin{enumerate}
    \item Probability of the outcome if I have bad habits (smoke and don’t exercise)? How about if I have good habits (don’t smoke and do exercise)?
  \end{enumerate}
\end{enumerate}

```{r q1q2, echo=FALSE, results='asis'}

## 1.
# Create the Bayesian network to analyze the survey results. It will be the product of 
# all the probabilities corresponding to the full joint distribution. 
risk_factors_bayes_net = create_bayes_net()

## 2.
# For each of the four health outcomes (diabetes, stroke, heart attack, angina), answer the following by 
# querying the network

## (a)
# What is the probability of the outcome if I have bad habits (smoke and don’t exercise)? 
outcome_tables = get_health_outcomes(bayes_net=risk_factors_bayes_net, 
                                     health_outcomes=health_outcomes, 
                                     observe_variables=habit_variables, 
                                     observe_values=bad_habit_values)
table_counter = 0
for (factor_table in outcome_tables) {
  table_counter = table_counter + 1
  caption_text = paste("Probability for ", health_outcomes[table_counter], " with smoking and no exercise")
  print(kable(factor_table, caption = caption_text))  
}

# How about if I have good habits (don’t smoke and do exercise)?
outcome_tables = get_health_outcomes(bayes_net=risk_factors_bayes_net, health_outcomes=health_outcomes, observe_variables=habit_variables, observe_values=good_habit_values)
table_counter = 0
for (factor_table in outcome_tables) {
  table_counter = table_counter + 1
  caption_text = paste("Probability for ", health_outcomes[table_counter], " with no smoking and exercise")
  print(kable(factor_table, caption = caption_text))  
}
```

\begin{enumerate}

  \item[(b)] What is the probability of the outcome if I have poor health (high blood pressure, high cholesterol, and overweight)? What if I have good health (low blood pressure, low cholesterol, and normal weight)?

\end{enumerate}

```{r q2contd, echo=FALSE, results='asis'}

## (b)
# What is the probability of the outcome if I have poor health (high blood pressure, high cholesterol, and overweight)?
outcome_tables = get_health_outcomes(bayes_net=risk_factors_bayes_net, health_outcomes=health_outcomes, observe_variables=health_variables, observe_values=bad_health_values)
table_counter = 0
for (factor_table in outcome_tables) {
  table_counter = table_counter + 1
  caption_text = paste("Probability for ", health_outcomes[table_counter], " with bad health")
  print(kable(factor_table, caption = caption_text))  
}

# What if I have good health (low blood pressure, low cholesterol, and normal weight)?
outcome_tables = get_health_outcomes(bayes_net=risk_factors_bayes_net, health_outcomes=health_outcomes, observe_variables=health_variables, observe_values=good_health_values)
table_counter = 0
for (factor_table in outcome_tables) {
  table_counter = table_counter + 1
  caption_text = paste("Probability for ", health_outcomes[table_counter], " with good health")
  print(kable(factor_table, caption = caption_text))  
}
```
